---
title: "EDA"
author: "Stephanie Hoglund"
output: html_document
date: "2023-09-29"
---

```{r setup, include=FALSE, echo=TRUE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(readr)
library(DataExplorer)
library(magrittr)
library(janitor)
library(plotly)
library(reticulate)
use_python("/Users/stephaniehoglund/Documents/Github/chef-it-up/.venv/bin/python")

`%!in%` <- Negate("%in%")
```

# As a chef's assistant at this cooking institution from which I got the data from, I have some background knowledge that helped shape this analysis and understand some nuances of the data.

# Initial Look
I'll start with seeing the data quality I have to scope out the data cleaning and manipulation I'll need to do. I like to use the `DataExplorer` package as a first step because of their simple functions that output nice visuals.
```{r}
fill_rates <- read_csv("./data/FillRateReport - 2023-06-15T205913.796.csv",
                       guess_max = 1000) %>% 
  clean_names(case = "none") %>% 
  select(-c(Store_ID, Store))

```


Using the visual below, it looks like I don't have too much missing data (<2%) and there's a 70-30 split between categorical and quantitative columns, respectively. 
<center>
```{r}
plot_intro(fill_rates)
```
</center>
<br>

Let's see what exactly is missing...
<center>
```{r}
plot_missing(fill_rates)
```
</center>
<br><br>
`Chef_Name` is missing ~23% of the time. Not too damning to remove from the dataset, so I'll keep it in for now. However, I noticed that pre-pandemic, the chef names are correctly assigned but post-pandemic, the chef name was either missing or filled in with the resident chef, thus hiding the true instructor's name. For this reason, I'll exclude the chef's name as a feature for the model.

Because a large amount of the columns is 'discrete', this makes me think that numerical data got read in as text. Let's see the first 10 records...
```{r}
fill_rates %>% head(10)
```

Sure enough, `Date` has the backward slashes and the last 4 columns have their units (%,\$) typed out which automatically makes that field a character type. We'll clean that now:

```{r}
fill_rates %<>% 
  mutate(Date = as.Date(Date, tryFormats = c("%m/%d/%Y")),
         Fill_Rate_percent = as.numeric(gsub("%", "", Fill_Rate_percent))) %>% 
  mutate_at(vars(Cost, Potential, Actual), ~as.numeric(gsub("[$,]", "", .)))

fill_rates %>% head(10)
```

Other notes:

* `SKU` seems like it could be a unique identifier for each class instance
* `Class_Name` and `Paid_in_Full`/`Fill_Rate_percent` indicate whether a class was cancelled
* Taking the delta between `Potential` and `Actual` tells you lost profit


# Summary Statistics
Now that the data has the proper types, we'll look into basic summary statistics.

```{r}
fill_rates %>% 
  summary()
```


As a chef's assistant, I know the max number of students we can have in a class is 16. The data confirms this rule. 

## Cost
Max Cost is the most surprising statistic to me from previous tables because \$2000/person seems steep. However, I think that private events and multi-day summer camps may be included in this dataset because they cost more than a typical class. Thus, they would need to be removed for future demand forecasting as they are outside the scope of this project.

```{r}
fill_rates %>% filter(Cost > 99) %>% 
  mutate(Class_Name_cleaned = gsub(" ?\\(.*\\)", "", Class_Name),
         Class_Name_cleaned = gsub(": .*$", "", Class_Name_cleaned)) %>% 
  group_by(Class_Name_cleaned) %>% 
  summarise(n = n(),
            Avg_Price = mean(Cost, na.rm = TRUE)) %>% 
  arrange(desc(n))
```

After cleaning up the class name a little bit to somewhat standardize the groupings and looking at classes with a unit cost over $99 (our regular cost), we can see that a majority of these classes are private events, kids/teens camps, and special classes that come with extra goodies ("free" pans or knives). 

I'm now curious to see if classes that come with goodies are filled more than the regular classes without goodies - I'll test this hypothesis at the end. I would also want to see which type of goodie entices people to take a class: do people find getting a casserole dish or a skillet worth their money? For me, I'll take the skillet because you can never have too many of those.

For this project, we'll remove the private events, camps, and classes with goodies. While I'm at it, I'll also remove the classes that said "cancelled".
```{r echo=TRUE}
fill_rates_classes <- fill_rates %>% 
  filter(!grepl("Private Event", Class_Name),
         !grepl("Series", Class_Name),
         !grepl("\\+", Class_Name),
         !grepl("cancelled", Class_Name))
```

Let's see what the spread of prices is now...
<center>
```{r}
fill_rates_classes %>% 
  plot_ly(type = "violin",
          y = ~Cost,
          box = list(visible = TRUE),
          meanline = list(visible = TRUE),
          x0 = "Cost per Class") %>% 
  layout(yaxis = list(title = ""))
```
</center>

Using the boxplot, the average price of a class hovers at \$78 with the lowest class at \$47 and the highest at \$109. I think the \$47 is for the family classes and the \$109 is for the seafood classes. 

Looking at the violin portion of the plot, a majority of the classes are priced at or above the mean. I wonder if this is because of inflation and prices increased over time due to rising food costs or the store felt they could increase profits this way. Either way I would want to see if fill rates still are impacted by this changing price.

```{r}
fill_rates_classes %>% 
  mutate(Year = year(Date),
         Month = month(Date),
         Month = ifelse(Month < 10, str_pad(Month, 2, "left", 0), Month),
         Year_Month = paste(Year, Month, sep = "-")) %>% 
  group_by(Year_Month) %>% 
  summarise(Avg_Cost = mean(Cost, na.rm = TRUE)) %>% 
  plot_ly(type = "scatter",
          mode = "lines+markers",
          x = ~Year_Month,
          y = ~Avg_Cost) %>% 
  layout(title = "Class Prices over Time",
         xaxis = list(title = "Date"),
         yaxis = list(title = "Average Cost per Class"))
```

The COVID-19 pandemic was pivotal movement, in more ways than one, but in this case with prices. Before the pandemic hit in March 2020, prices in mid-2019 were hovering just below \$65 before being raised \$10 to close the year out. Excluding the large dip in prices during the pandemic, we see that prices did not go back to pre-pandemic levels. Instead, they increased another \$10 in December 2020 and it hovers at \$77 for another year before seeing yet another price increase in Summer 2022. Not sure if the company was trying to make up for lost profits during the pandemic or if they're taking advantage of the American consumer suddenly buying after being locked down. I know this particular company had to close many of their stores around the country and my store had condensed traffic from two other stores in the area. So I think it was a combination of all these factors, wanting to make as much profit as possible, inflation, and rising food costs that led us to where we are now with an average class costing \$87.


## Time of Day
From my experience, there's a difference between morning and evening classes not only with the menu but also class capacity. The morning classes don't experience the same traffic and volume as the evening classes because many people have jobs during the day. I'll classify the classes into morning, afternoon, and evening buckets, then test the hypothesis that the average fill rate for morning classes is less than evening.

```{r}
fill_rates_classes %<>% 
  mutate(Hour = hour(Time),
         time_of_day = case_when(
           Hour < 13 ~ "Morning",
           Hour < 17 ~ "Afternoon",
           TRUE ~ "Evening"
         ))

fill_rates_classes$time_of_day <- factor(fill_rates_classes$time_of_day)
```

Does the store offer the same number of morning, afternoon, and evening classes?
```{r}
fill_rates_classes %>% 
  count(time_of_day)
```

Overall, yes, the store does seem to offer the same amount of classes throughout the day. 


What is the spread of classes during the day? Aka how filled are classes throughout the day? Are they similar?
<center>
```{r}
fig <- ggplot(fill_rates_classes,
       aes(x = Fill_Rate_percent, 
           fill = time_of_day)) +
  geom_density(alpha = 0.2)

ggplotly(fig) %>% 
  layout(xaxis = list(title = "Fill Rate (%)"),
         legend = list(title = list(text = "<b>Time of Day</b>")))
```
</center>

The evening and afternoon classes look to be filled more often than the morning classes... but is there a significant difference? Check out the last section of this report.



## Date Range
The data spans exactly 4 years. Let's see how many classes were on the schedule during this time versus how many were actually filled.

Here's are simple histograms based on the dates. I included all the classes and then the ones that were scheduled (after removing the "cancelled" classes from the Cost section). As you can see, there's a dip in scheduled classes between July 2020 and January 2021 because of the COVID-19 pandemic. Many stores, including this cooking class one, were shut down so I believe scheduling was reduced. We'll also have to remove additional records between March 11, 2020, when the World Health Organization (WHO) declared the pandemic [according to the CDC](https://www.cdc.gov/museum/timeline/covid19.html), and November 24, 2021 when this store stopped reducing the class capacity, to use transactions during "regular" operations for demand forecasting. The store took safety precautions and cut class capacity in half (from 16 to 8 students) during the pandemic.

<center>
```{r}
plot_ly(alpha = 0.6) %>% 
  add_histogram(x = ~fill_rates$Date,
                name = "All Classes") %>% 
  add_histogram(x = ~fill_rates_classes$Date,
                name = "Scheduled Classes") %>% 
  layout(title = "Date Range",
         barmode = "overlay")
```
</center>

I found out the last date of reduced class capacity by:
```{r echo=TRUE}
fill_rates %>% 
  filter(Max_Seats == 8) %>% 
  slice_max(Date) %>% 
  pull(Date)
```


For this reason, I'll remove records in the previously-mentioned time frame (March 11, 2020 - November 24, 2021).
<center>
```{r}
fill_rates_nocovid <- fill_rates_classes %>% 
  filter(Date < "2020-03-11" |
         Date > "2021-11-24")

fill_rates_nocovid %>% 
  plot_ly(type = "histogram",
          x = ~Date) %>% 
  layout(title = "Date Range (Scheduled Non-COVID)")
```
</center>


```{r}
reduced_percent <- (nrow(fill_rates_nocovid) - nrow(fill_rates_classes))/nrow(fill_rates) * 100
```

Filtering out the COVID-19 records reduced our dataset by **`r abs(round(reduced_percent, 1))`%**.


### Weekday vs Weekend Classes
Fill rates may be impacted if the class is offered on a weekday or weekend because people typically work corporate jobs during the week. For this analysis, I'll classify Monday through Thursday as a "weekday" and Friday through Sunday as a "weekend" by creating a binary variable, `is_weekend`, for my future model.

```{r}
fill_rates_nocovid %<>% 
  mutate(day_of_week = factor(wday(Date, label=TRUE)),
         is_weekend = factor(ifelse(day_of_week %in% c('Fri', 'Sat', 'Sun'), 1, 0), labels = c("Weekday", "Weekend")))
```


What is the distribution of fill rates for a weekend versus a weekday?
<center>
```{r}
fill_rates_nocovid %>% 
  plot_ly(type = "box",
          x = ~is_weekend,
          y = ~Fill_Rate_percent,
          color = ~is_weekend,
          boxmean = TRUE) %>% 
  layout(title = "Fill Rate for Weekday vs Weekend",
         xaxis = list(title = ""),
         yaxis = list(title = "Fill Rate (%)"))
```
</center>

```{r}
fill_rates_nocovid %>% 
  group_by(is_weekend) %>% 
  summarise(Count = n(),
            Min = min(Fill_Rate_percent, na.rm = TRUE),
            Q1 = quantile(Fill_Rate_percent, 0.25, na.rm = TRUE),
            Mean = mean(Fill_Rate_percent, na.rm = TRUE),
            Median = median(Fill_Rate_percent, na.rm = TRUE),
            Q3 = quantile(Fill_Rate_percent, 0.75, na.rm = TRUE),
            Max = max(Fill_Rate_percent, na.rm = TRUE),
            SD = sd(Fill_Rate_percent, na.rm = TRUE))
```


The average fill rate for the Weekend is greater than the Weekday, which was expected. 
<br>
What's interesting to see is that, for the Weekday, the mean is above the median - this finding suggests the data is *positively skewed* because extremely high values are pulling the mean to the right while a majority of the values hover around 43.75. For the Weekend, we see the opposite, the mean is below the median, which suggests the data is *negatively skewed* because extremely low values are pulling the mean to the left while a majority of the values hover around 78.125. 
<br>
It's also crazy to see Weekend Q1 at 50, which means that 25% of the Weekend classes have fill rates below half or that 75% of the time Weekend classes are at least half full.


## Holiday Season
Another big factor I've seen with cooking classes aside from price and time of day/week is how close we are to the holidays. The holidays are busy and people love to schedule an event when family and friends come into town, especially if it's snowing outside (I kinda stole my own thunder here, but we'll address weather in the next section - pun intended). 

```{python}
from datetime import datetime, date
import pandas as pd
import holidays

us_holidays = holidays.US(expand=True)

fill_rates_nocovid = r.fill_rates_nocovid

# add column if class is on a holiday
fill_rates_nocovid['is_holiday'] = fill_rates_nocovid['Date'].apply(lambda x: x in us_holidays)

# add column for how far away a class is from a holiday (either before or after the holiday)
fill_rates_nocovid['closest_holiday'] = fill_rates_nocovid['Date'].apply(lambda x: min(us_holidays, key=lambda y: abs(y - x)))

fill_rates_nocovid['closest_holiday_name'] = fill_rates_nocovid['closest_holiday'].apply(lambda x: us_holidays.get(x))

fill_rates_nocovid[['Date', 'closest_holiday']] = fill_rates_nocovid[['Date', 'closest_holiday']].apply(pd.to_datetime)

fill_rates_nocovid['proximity_to_holiday'] = abs(fill_rates_nocovid['Date'] - fill_rates_nocovid['closest_holiday']).dt.days

```

```{r}
fill_rates_nocovid <- py$fill_rates_nocovid %>% 
  mutate(closest_holiday = as.Date(closest_holiday))


fill_rates_nocovid %>% 
  plot_ly(type = "scatter",
          x = ~proximity_to_holiday,
          y = ~Fill_Rate_percent) %>% 
  layout(title = "Fill Rates Near the Holidays",
         xaxis = list(title ="Days to a Holiday"),
         yaxis = list(title = "Fill Rate (%)"))
```


From the above plot, there doesn't seem to be a linear relationship between holiday proximity and fill rate percentage. To further confirm this conclusion with statistics, the Pearson and Spearman correlation coefficients both hover at `r round(cor(fill_rates_nocovid$Fill_Rate_percent, fill_rates_nocovid$proximity_to_holiday), 3)` and `r round(cor(fill_rates_nocovid$Fill_Rate_percent, fill_rates_nocovid$proximity_to_holiday, method = "spearman"), 3)`, respectively.


For my curiosity, which holiday brought in more people (i.e. higher fill rates)?
```{r}
fill_rates_nocovid %>% 
  mutate(closest_holiday_name = gsub(" \\(Observed\\)","", closest_holiday_name)) %>% 
  group_by(closest_holiday_name) %>% 
  summarise(`Num Classes Offered` = n(),
            `Avg Fill Rate` = mean(Fill_Rate_percent, na.rm = TRUE)) %>% 
  arrange(desc(`Avg Fill Rate`))
```

Although the most classes were offered around Washington's Birthday in mid-February, classes around New Year's were the most filled with Washington's Birthday a close second again. This finding checks out because we tried to schedule classes during school breaks when people wouldn't necessarily be traveling out of town (like for Thanksgiving or Christmas), but still wanted to do something. 


## Weather
The final feature I wanted to include in this analysis is weather. Weather impacts cooking classes significantly because we would always have people arrive late or cancel completely if there was significant snow or rain. The temperature also influenced fill rates because if it was too hot or too cold outside, people would rush inside for an air-conditioned activity or for the warmth of food. But if the temperature was just right, our store would be a ghost town - I called this the Goldilocks Effect.

I'm going to pull in data from the National Oceanic and Atmospheric Administration (NOAA) for the region and add average temperature and whether it snowed/rained or not as new features to evaluate for my model.

```{r}

```





# Other Analyses
If I had the data, I would want to tie in retail data to see which classes bring in more revenue from the students shopping during the class. It's protocol to give the students a "break", but really we want them to shop more in the store while the kitchen assistants can catch up on cleaning.

## Are special Classes with goodies filled more than classes without goodies?
Special classes that came with a "free" gift are denoted by a "+" in the class name and not followed by "More" or "Dessert". Let's use a **one-way t-test** to compare the fill rate means of two groups: classes with and without goodies. I assume the groups have unequal variances because they have different sample sizes and their boxplots show different ranges of fill rate values.

* H0: there is no difference between the groups
* Ha: classes with goodies are filled more than classes without goodies

```{r}
special_classes <- fill_rates %>% 
  filter(grepl("\\+ \\w+", Class_Name),
         !grepl("cancelled", Class_Name),
         !grepl("\\+ (?=More|Dessert)", Class_Name, perl = TRUE)) %>% 
  mutate(Group = "Special")

# need to make sure we exclude the private and/or cancelled classes
non_special_classes <- fill_rates_classes %>% 
  mutate(Group = "Non-Special")

classes <- bind_rows(special_classes, non_special_classes) %>% 
  select(Group, Fill_Rate_percent)

classes %>% 
  plot_ly(type = "box",
          x = ~Group,
          y = ~Fill_Rate_percent,
          boxmean = TRUE)
```

```{r}
alpha <- 0.05

t_test <- t.test(Fill_Rate_percent ~ Group, 
       data = classes,
       alternative = "less",
       conf.level = 1-alpha)

t_test

if(t_test$p.value < alpha){
  print("Reject H0")
} else{
  print("Fail to reject H0")
}
```

Because the p-value is more than the alpha value of **0.05**, we *fail to reject the null hypothesis* and the test suggests that the mean fill rate for special classes is no different than that for non-special classes. However, if we set the alpha to **0.10**, we *reject the null hypothesis* and the test would suggest that the mean fill rate for special classes is greater than that for non-special classes. The alpha value is also known as the type I error rate, the probability that you reject the null hypothesis when the null hypothesis is true. However, I'll keep the significance level at 0.05 because of the different sample sizes in this test and keeping the significance level low shows strong support for the null hypothesis.



## Class Prices vs Fill Rates
In the Cost section earlier, we saw how the company increased prices steadily after the pandemic. Here, we'll categorize classes into price buckets and see which bucket is filled more often. I'll use the following buckets:

* Group 1: <\$70
* Group 2: \$70-80
* Group 3: \$80<

```{r}
class_rates <- fill_rates_nocovid %>% 
  mutate(Price_Group = case_when(
    Cost < 70 ~ "<$70",
    Cost < 80 ~ "$70-80",
    TRUE ~ "$80+"
  )) %>% 
  select(Price_Group, Fill_Rate_percent)


class_rates %>% 
  plot_ly(type = "box",
          x = ~Price_Group,
          y = ~Fill_Rate_percent,
          boxmean = TRUE)
```


Using Analysis of Variance (ANOVA), I'll test to see which price group has a statistically significant higher mean fill rate.

H0: there is no difference in mean fill rates between all 3 price groups
Ha: at least one price group differs in mean fill rate

```{r}
price_anova <- aov(Fill_Rate_percent ~ Price_Group, class_rates)

summary(price_anova)
```

Using a significance level of 0.05, we fail to reject the null hypothesis (just barely) and agree that there is no difference in mean fill rates for the price groups. Therefore, the changes in pricing are not stopping the American consumer from booking cooking classes. Our testing stops here and we don't need to continue further with Tukey's Procedure and finding which price group is statistically significant.

## Average fill rate for morning classes is less than evening
Similar to the previous section with the price groups, we'll analyze here if the time at which the class is offered affects the class's fill rate.

```{r}
fill_rates_nocovid %>% 
  plot_ly(type = "box",
          x = ~time_of_day,
          y = ~Fill_Rate_percent,
          boxmean = TRUE)
```

Using ANOVA, let's see what the results are at an alpha level of 0.05:
```{r}
tod_anova <- aov(Fill_Rate_percent ~ time_of_day, fill_rates_nocovid)

summary(tod_anova)
```

The ANOVA test shows that we reject the null hypothesis because there's strong evidence to support that at least one group has a different mean fill rate. Let's use Tukey's procedure to see which group(s) is/are statistically different from each other.

```{r}
TukeyHSD(tod_anova)
```

```{r}
plot(TukeyHSD(tod_anova))
```

Using the table and plot, we see that all groups are statistically different from each other because all the p-values are less than alpha of 0.05 (so we reject evey null hypothesis of the groups being the same) and none of the confidence intervals include zero.

